### 2016-01-20 writeups/msr_workshop/paper.tex
Initial commit

\documentclass[a4paper,twocolumn]{article}

\title{
  It's Never Too Late To Learn\\
  \emph{A case for ML in distributed systems}
}
\author{}
\date{}

\begin{document}
\maketitle

\section{Motivation}

Building a modern distributed service that guarantees service-level objectives
is hard. Complexity is introduced by the deployment environment and exacerbated
as the service evolves with time to accommodate hardware/environment changes or
support new features.
For example, to build a simple caching layer that guarantees a certain response
latency and throughput, developers need to take into account the number of
machines available to run that service, the network connecting these machines,
the storage hierarchy within each machine, and the expected workload that the
caching layer will be servicing. A change in any one of these parameters, could
render the service inefficient.

Traditionally, this complexity is addressed with custom-built systems that
are optimized for specific configurations and objective functions. Is there a
general, principled, approach to dealing with complexity in the design of
distributed systems?
In this position paper, we propose a research project to study the use of
Machine Learning (ML) in the design of distributed systems. We highlight
specific examples of where ML can help, and discuss potential issues that need
to be addressed.

\section{Examples}

\subsection{Known unknowns}
Systems often make decisions without having complete knowledge of all the
factors that affect, or could be affected by, these decisions.  For example, a
caching system evicts items based on what it thinks future requests will look
like, even though it cannot actually know what the future will look like. Here,
decisions are made in anticipation of future events.  In other cases, this
incomplete knowledge is due to the absence of data-computation locality. For
example, a load balancer makes decisions based on what it thinks is the current
load on other machines rather than what it actually is.

In both of these situations, assumptions are made about the unknowns in
\emph{design-time}--resulting in different different routing and caching
algorithms. A change to the \emph{run-time} environment could render these
decisions wildly inefficient. By deferring to an ML agent, these decisions can
be made online, and can be nimble in adapting to any changes to the run-time
environment.

\subsection{Unknown unknowns}
In some situations, system designers have no way of knowing deployment details
that would impact the system's performance. For example, a system could be built
for external deployment and used by third party customers, or it could be
deployed in a multi-tenant environment where co-located services could impact
its performance.

In these cases, it is near impossible for developers to account for all the
unknown variables in design-time. However, an ML agent is able to make decisions
online based on inputs from the run-time environment.

\section{Issues of ML in Systems}

In this section, we list preliminary issues and open questions that might be
problematic for ML in systems and are worth studying.

\subsection{Hierarchy}
Distributed systems are often designed in a hierarchical manner, where
subcomponents are monitoring and managing others. These components and layers of
indirection simplify reasoning about the system at large. However, it is not
clear how will hierarchy affect an ML agent.
\begin{itemize}
  \item Will hierarchy aid the learning system via parallelism? Or will it hinder
    it since not all features are visible at every level?
  \item Should learning happen in each component of the system? Or should it be
    centralized?
  \item If learning is happening at multiple levels of the hierarchy, will that
    create interference between the learners? Or will they compliment each
    other?
\end{itemize}

\subsection{Decisions vs enforcement}
Decisions made by an ML agent might be executed by different components and
machines. How can the decisions be executed consistently, and reward signals
tracked correctly at large scale?

\subsection{Feature selection}
What features of the run-time system are important to the ML agent? How can they
be captured effectively? Can the ML agent learn from incomplete data? For
example, if an ML agent needs to make decisions regarding performance
optimization, would ignoring garbage collection activities or the VMM result in
corrupting the learned model? How sensitive should the learner be?

\subsection{Online vs offline}
An offline ML agent is trained with pre-labeled data. However, is this possible
in the context of distributed systems? An online learning systems is perhaps
more logical in this context, however, how should the reward signals be set up?
How do we quantify the costs of exploration vs exploitation of learned
strategies?

\end{document}

### 2016-01-21 writeups/msr_workshop/paper.tex
microedits

@@
 machines available to run that service, the network connecting these machines,
-the storage hierarchy within each machine, and the expected workload that the
-caching layer will be servicing. A change in any one of these parameters, could
-render the service inefficient.
+the storage hierarchy within each machine, the expected workload that the
+caching layer will be servicing, and interference from other applications on the network and machines.
+A change in any one of these parameters could render the service inefficient.
 
@@
 are optimized for specific configurations and objective functions. Is there a
-general, principled, approach to dealing with complexity in the design of
+general, principled approach to dealing with complexity in the design of
 distributed systems?
@@
 indirection simplify reasoning about the system at large. However, it is not
-clear how will hierarchy affect an ML agent.
+clear how hierarchy will affect an ML agent.
 \begin{itemize}
@@
   \item If learning is happening at multiple levels of the hierarchy, will that
-    create interference between the learners? Or will they compliment each
+    create interference between the learners? Or will they complement each
     other?

### 2016-01-22 writeups/msr_workshop/paper.tex
More changes after conversation with Mahesh and Sid

@@
   It's Never Too Late To Learn\\
-  \emph{A case for ML in distributed systems}
+  \emph{Using Online Machine-Learning in Distributed Systems}
 }
@@
 the storage hierarchy within each machine, the expected workload that the
-caching layer will be servicing, and interference from other applications on the network and machines.
-A change in any one of these parameters could render the service inefficient.
+caching layer will be servicing, and interference from other applications on the
+network and machines.  A change in any one of these parameters could render the
+service wildly inefficient.
 
-Traditionally, this complexity is addressed with custom-built systems that
-are optimized for specific configurations and objective functions. Is there a
-general, principled approach to dealing with complexity in the design of
-distributed systems?
-In this position paper, we propose a research project to study the use of
-Machine Learning (ML) in the design of distributed systems. We highlight
-specific examples of where ML can help, and discuss potential issues that need
-to be addressed.
+Traditionally, this complexity is addressed by using highly-skilled developers
+to build custom systems that are optimized for specific configurations and
+objective functions.  Is there a general, principled approach to dealing with
+complexity in distributed systems?
+In this position paper, we propose a research project to study the use of Online
+Machine Learning (ML) in distributed systems.  Hand-tuning the design and
+configuration of distributed systems does not scale well with increased
+complexity of software design, hardware architecture, deployment environment,
+and strict service-level objectives.
+To highlight this, here are some examples of where we think ML can help:
 
-\section{Examples}
-
-\subsection{Known unknowns}
+\subsection*{Known unknowns}
 Systems often make decisions without having complete knowledge of all the
@@
 
-In both of these situations, assumptions are made about the unknowns in
-\emph{design-time}--resulting in different different routing and caching
+In both of these situations, assumptions are made about the unknowns offline in
+\emph{design-time}--resulting in different routing and caching
 algorithms. A change to the \emph{run-time} environment could render these
@@
 
-\subsection{Unknown unknowns}
+\subsection*{Unknown unknowns}
 In some situations, system designers have no way of knowing deployment details
@@
 
+To be clear, ML has been used in some distributed systems. However, it has not
+been used in a general way to make online decisions about core distributed
+systems implementation details. We advocate that developers should focus on
+high-level objectives of their system, and use online ML to make to compute the
+specifics (such as policies for request routing, replica placement, or failure
+monitoring) at run-time based on observed metrics.
+
 \section{Issues of ML in Systems}
 
-In this section, we list preliminary issues and open questions that might be
-problematic for ML in systems and are worth studying.
+Using online ML in distributed systems poses many issues and open questions.
+Here are some preliminary questions that we think are worth studying:
 
-\subsection{Hierarchy}
+\subsection*{Framing}
+An online ML loop, like Microsoft's Multi-World Testing Decision Service, views
+the external world in terms of \emph{context} (a summarization of the world's
+history), \emph{actions} suggested by the agent, and \emph{rewards} collected on
+these actions. How can a distributed system's current state and service-level
+objectives be translated into context and rewards? In some scenarios, such as
+request routing for load balancing, the translation is easy. In others, like
+replica placement, the translation is hard.
+
+\subsection*{Hierarchy}
 Distributed systems are often designed in a hierarchical manner, where
@@
 
-\subsection{Decisions vs enforcement}
-Decisions made by an ML agent might be executed by different components and
-machines. How can the decisions be executed consistently, and reward signals
-tracked correctly at large scale?
+\subsection*{Locality}
+Decisions made by an ML agent in distributed system are likely to rely on input
+data collected from diverse and physically distributed components. As a result,
+the input to the learning loop might be stale or inconsistent. What could be
+done to mitigate those issues while making reasonable real-time decisions?
+Additionally, the decisions outputted by the learning agent need to be executed
+by diverse and distributed components. How could routine distributed systems
+issues such failures and delays in an asynchronous setting affect the learning
+system?
 
-\subsection{Feature selection}
+\subsection*{Feature selection}
 What features of the run-time system are important to the ML agent? How can they
@@
 
-\subsection{Online vs offline}
-An offline ML agent is trained with pre-labeled data. However, is this possible
-in the context of distributed systems? An online learning systems is perhaps
-more logical in this context, however, how should the reward signals be set up?
-How do we quantify the costs of exploration vs exploitation of learned
-strategies?
+\subsection*{Training}
+An ML agent can be trained offline with pre-labeled data. However, is this
+possible in the context of distributed systems? An online learning systems is
+more logical in this context, however, how do we quantify the costs of
+exploration vs exploitation of learned strategies?
 

### 2016-01-22 writeups/msr_workshop/paper.tex
notes on what to integrate/add

@@
   It's Never Too Late To Learn\\
-  \emph{Using Online Machine-Learning in Distributed Systems}
+  \emph{Using Online Machine Learning in Distributed Systems}
 }
@@
 To highlight this, here are some examples of where we think ML can help:
+
+- contextual multi-armed bandits. idea is to provide a context, get an action,
+and only observe reward for that action. This matches the dist systems world well. It will generate new
+policies using ML that we could not come up with by just thinking, and may not
+even be able to describe. We argue in favor of opacity and rely instead on the guarantees
+provided by the ML optimization.
+
+We describe some of hte unkowns we face and where ML can come in
 
@@
 
-\section{Issues of ML in Systems}
+\section{Applying Online Learning to Distributed Systems}
+
+- Framework is to threat ML as a black box. The Decision Service is an exmaple of such a system.
+The blackbox takes as input a context, teslls us an action, and then expects us to give it
+a reward. Thus it can be placed anywher a decision has to be made, such as which request
+to route to, where to place replicas, etc.. Despite the simplicity provided by the ML interface
+and we assume the rest of it is solved: it colelcts the data, optimizes a model, and delivers the model
+to us quickly (let's even assume the latency of htis loop, which can be minutes, is good enough to adapt
+to the rate of change in our environment). Now the remaining piece is to figure out what
+to enter as context, how to execut the actions, and how to determine or report rewards.
+
+We argue that there are big challenges even in doing just these things...
 
@@
 Here are some preliminary questions that we think are worth studying:
+- Questions viewed from the lens of characteristics of distributed systems
+
+[merge below above]
 
@@
 
+- Should feature vectors cut across the levels or not?
+- In general should be composable if at each level can define decision and reward,
+and exploration randomization is different at each level
+
 \subsection*{Locality}
@@
 
+- quesiton of what features to include is there, but being conservative usually the right
+way to go
+- bigger quesiton is how ml component deals with omissions and other problems. Separate
+into failure sections? Yeah, replace this with ``failures'' and move feature selection stuff
+down to last section
+
 \subsection*{Training}
@@
 
+- Probably omit this, or replace with ``Randomization''. Basically, we already know
+that exploratoin-exploitation is the right approach, and that will be encapsulated
+by the ML component (its decisions will reflect this, and it will take care of
+computing probabilities etc.). But there's an interesting point about inherent randomization
+in systems
+
+\section{Expanding the scope}
+
+Two drawbacks: opacity and scope.
+
+Opacity (also, which features matter?)
+- We are also pursuing work on cracking models open to see what they are doing inside.
+There are different ways to gain insight as to what a model is doing. For example a simple
+linear model may learn a vector of weights on the supplied features. These weights can be
+studied, and indeed some techniques such as lasso are oftne used to reduce the number of features. However, these
+techniques need to be handled carefully. Just becasue a feature has 0 weight doesn't mean it
+can be removed: perhaps it is unimportant preceisely because it exists, and its removal all of
+asudden changes the weight on other features.
+
+Scope:
+The above replaces the policy behind a decision with one backed by an ML learning system.
+Indeed systems like hte DS were designed to intervene at the decision making point in a
+slim way. The above outlines the issues which tackled would result in ML based policies,
+new ones. However, once the above has been mastered, there is the question of whether ML
+can influence distributed systems design at a higher level. For example, can it suggest
+different mechanisms for achieving a goal, or perhaps a different design of the system in the first place?
+
 \end{document}

### 2016-01-22 writeups/msr_workshop/paper.tex
new intro

@@
 \documentclass[a4paper,twocolumn]{article}
+
+\newcommand{\ignore}[1]{}
 
@@
 support new features.
-For example, to build a simple caching layer that guarantees a certain response
-latency and throughput, developers need to take into account the number of
-machines available to run that service, the network connecting these machines,
-the storage hierarchy within each machine, the expected workload that the
-caching layer will be servicing, and interference from other applications on the
-network and machines.  A change in any one of these parameters could render the
-service wildly inefficient.
+For example, to build a simple caching layer that guarantees a certain hit rate,
+response latency, and throughput, developers need to take into account the expected
+workload hitting the cache, the number of machines available to run it, the network
+connecting these machines, the storage hierarchy within each machine, and interference
+from other applications running on the network and machines.  A change in any one
+of these factors could render the service inefficient.
 
@@
 to build custom systems that are optimized for specific configurations and
-objective functions.  Is there a general, principled approach to dealing with
-complexity in distributed systems?
-In this position paper, we propose a research project to study the use of Online
-Machine Learning (ML) in distributed systems.  Hand-tuning the design and
-configuration of distributed systems does not scale well with increased
-complexity of software design, hardware architecture, deployment environment,
-and strict service-level objectives.
-To highlight this, here are some examples of where we think ML can help:
+objective functions.  Is there a general, principled approach for dealing with
+complexity in distributed systems? Hand-tuning the design and configuration of a system
+does not scale well with increased complexity and changes in software design, hardware
+architecture, deployment environment, and service-level objectives.
 
-- contextual multi-armed bandits. idea is to provide a context, get an action,
-and only observe reward for that action. This matches the dist systems world well. It will generate new
-policies using ML that we could not come up with by just thinking, and may not
-even be able to describe. We argue in favor of opacity and rely instead on the guarantees
-provided by the ML optimization.
+We propose to use online learning to cope with this complexity. Specifically, we
+view a {\em decision} made by a distributed system in the framework of
+contextual online learning with partial feedback. This is a natural fit: we typically
+have plenty of context surrounding a decision (which we may not know how to use), the decision
+is made online, and we only receive feedback for the decision that was made. Thus, we can
+replace a hand-designed policy with a new policy that has been optimized by a contextual
+learning system. By supplying rich contexts to this policy, we can capture the complexity of
+our environment without having to understand it; by optimizing the policy continuously, we can
+can adapt to changes in the environment or service over time.
 
-We describe some of hte unkowns we face and where ML can come in
+Machine learning has certainly been used in the past to optimize system decisions. Most
+of this use has been restricted to offline settings with full information: for example,
+supervised learning is often used to train models on data that has been annotated with
+correct labels. Some systems work in an online setting with full information: for example,
+a cache that applies an experts learning framework to dynamically switch between existing
+(hand-designed) cache eviction policies, by running all policies in parallel and measuring
+their performance. Systems that work in an online setting with partial information tend to
+be for news or ad recommendation, but examples from systems infrastructure exist: for example,
+caching dynamic query content [], tuning parameters of multicore data structures [], allocating
+resources to cluster computing tasks [], and allocating servers to web applications [].
+In all of these systems, the context used to make a decision, and the
+decision itself, are {\em local} to a machine. This is achieved by either devising a decentralized
+algorithm in the first place, or giving up on global optimality. Thus we are not aware of any
+system that makes ``global'' decisions based on truly distributed architecture or state.
+
+And yet we use hand-designed policies to make these decisions all the time, such as for request
+routing, replica placement, replica selection, cloud resource allocation, failure recovery tasks, 
+and others. What is it about distributed systems that makes it difficult to apply online learning
+at standard decision points? This proposal identifies fundamental challenges posed by distributed
+systems that must be analyzed and addressed in order to make online learning a standard tool for
+cloud optimization.  These challenges hold even if we assume that the online learning system is
+a black-box that delivers optimal decisions in real-time--in other words, they are fundamental
+to the way we design distributed systems.  We begin by highlighting some of the unknown factors
+that contribute to these challenges, and then list the challenges themselves.
 
@@
 
+\ignore{
 To be clear, ML has been used in some distributed systems. However, it has not
@@
 monitoring) at run-time based on observed metrics.
+}
 
@@
 
+\subsection*{Locality}
+Decisions made by an ML agent in distributed system are likely to rely on input
+data collected from diverse and physically distributed components. As a result,
+the input to the learning loop might be stale or inconsistent. What could be
+done to mitigate those issues while making reasonable real-time decisions?
+Additionally, the decisions outputted by the learning agent need to be executed
+by diverse and distributed components. How could routine distributed systems
+issues such failures and delays in an asynchronous setting affect the learning
+system?
+
 \subsection*{Hierarchy}
@@
 and exploration randomization is different at each level
-
-\subsection*{Locality}
-Decisions made by an ML agent in distributed system are likely to rely on input
-data collected from diverse and physically distributed components. As a result,
-the input to the learning loop might be stale or inconsistent. What could be
-done to mitigate those issues while making reasonable real-time decisions?
-Additionally, the decisions outputted by the learning agent need to be executed
-by diverse and distributed components. How could routine distributed systems
-issues such failures and delays in an asynchronous setting affect the learning
-system?
 
@@
 
+\ignore{
+contextual multi-armed bandits. idea is to provide a context, get an action,
+and only observe reward for that action. This matches the dist systems world well. It will generate new
+policies using ML that we could not come up with by just thinking, and may not
+even be able to describe. We argue in favor of opacity and rely instead on the guarantees
+provided by the ML optimization.
+}
+
 Scope:

### 2016-01-22 writeups/msr_workshop/paper.tex
some edits

@@
 \subsection*{Known unknowns}
-Systems often make decisions without having complete knowledge of all the
-factors that affect, or could be affected by, these decisions.  For example, a
-caching system evicts items based on what it thinks future requests will look
-like, even though it cannot actually know what the future will look like. Here,
-decisions are made in anticipation of future events.  In other cases, this
-incomplete knowledge is due to the absence of data-computation locality. For
-example, a load balancer makes decisions based on what it thinks is the current
-load on other machines rather than what it actually is.
 
-In both of these situations, assumptions are made about the unknowns offline in
-\emph{design-time}--resulting in different routing and caching
-algorithms. A change to the \emph{run-time} environment could render these
-decisions wildly inefficient. By deferring to an ML agent, these decisions can
-be made online, and can be nimble in adapting to any changes to the run-time
-environment.
+Systems often make decisions without complete knowledge of the factors that affect,
+or could affect, those decisions.  For example, a caching system evicts items based
+on what it believes future requests will look like, without actually knowing the future.
+In other cases, incomplete knowledge is due to the absence of data-computation locality.
+For example, a load balancer makes decisions based on what it thinks the current load on
+machines is, e.g. based on data it collected recently, rather than the actual load at that
+very instant in time.
+
+In both of these situations, assumptions are made about the unknowns offline during
+\emph{design time}, resulting in different routing and caching algorithms. A change to
+the \emph{run-time} environment could easily render these choices suboptimal.
+
+\ignore{Online learning has the potential to make these decisions online, based on current contexts,
+and can be nimble in adapting to any changes to the run-time
+environment.}
 

### 2016-01-22 writeups/msr_workshop/paper.tex
smaller margins
work on challenges section

@@
 cloud optimization.  These challenges hold even if we assume that the online learning system is
-a black-box that delivers optimal decisions in real-time--in other words, they are fundamental
-to the way we design distributed systems.  We begin by highlighting some of the unknown factors
+a black-box that delivers optimal decisions with zero latency--in other words,
+they are fundamental to the way we design distributed systems.  We begin by highlighting some of the unknown factors
 that contribute to these challenges, and then list the challenges themselves.
@@
 
-\ignore{Online learning has the potential to make these decisions online, based on current contexts,
+\ignore{
+Online learning has the potential to make these decisions online, based on current contexts,
 and can be nimble in adapting to any changes to the run-time
-environment.}
+environment.
+}
 
 \subsection*{Unknown unknowns}
+
 In some situations, system designers have no way of knowing deployment details
-that would impact the system's performance. For example, a system could be built
-for external deployment and used by third party customers, or it could be
-deployed in a multi-tenant environment where co-located services could impact
-its performance.
+that might impact a system's performance. For example, a system could be
+built for external deployment and used by third-party customers, or it could be
+deployed in a multi-tenant environment where co-located services could
+affect its performance in unexpected ways.
 
-In these cases, it is near impossible for developers to account for all the
-unknown variables in design-time. However, an ML agent is able to make decisions
-online based on inputs from the run-time environment.
+In these cases, it is nearly impossible for developers to account for all 
+unknown variables at design time.
 
+\ignore{
+However, an ML agent is able to make
+decisions online based on inputs from the run-time environment.
+}
 \ignore{
@@
 
-\section{Applying Online Learning to Distributed Systems}
+\section{Using Online Learning in Distributed Systems}
 
-- Framework is to threat ML as a black box. The Decision Service is an exmaple of such a system.
-The blackbox takes as input a context, teslls us an action, and then expects us to give it
-a reward. Thus it can be placed anywher a decision has to be made, such as which request
-to route to, where to place replicas, etc.. Despite the simplicity provided by the ML interface
-and we assume the rest of it is solved: it colelcts the data, optimizes a model, and delivers the model
-to us quickly (let's even assume the latency of htis loop, which can be minutes, is good enough to adapt
-to the rate of change in our environment). Now the remaining piece is to figure out what
-to enter as context, how to execut the actions, and how to determine or report rewards.
+As mentioned earlier, we adopt the framework of contextual online learning
+with partial feedback.  To separate the systems concerns we are trying to
+highlight from machine learning concerns, we abstract the learning system as a
+black-box and assume it provides optimal decisions with zero latency. For a
+given decision over a discrete action space, the interface to the learning
+system is as follows: given a {\em context} that summarizes the state of the
+world, the learning system chooses an {\em action} to take, which our system
+does and later reports a {\em reward} indicating how good the action was. An
+example of a learning system that supports this interface is the Decision
+Service~\footnote{http://aka.ms/mwt}. An example of a distributed systems
+decision is routing requests at a load balancer: the context could be load
+information of each machine in the cluster, the action could be one of the
+possible machines, and the reward could be the latency of processin the request.
 
-We argue that there are big challenges even in doing just these things...
-
-Using online ML in distributed systems poses many issues and open questions.
-Here are some preliminary questions that we think are worth studying:
-- Questions viewed from the lens of characteristics of distributed systems
-
-[merge below above]
-
-\subsection*{Framing}
-An online ML loop, like Microsoft's Multi-World Testing Decision Service, views
-the external world in terms of \emph{context} (a summarization of the world's
-history), \emph{actions} suggested by the agent, and \emph{rewards} collected on
-these actions. How can a distributed system's current state and service-level
-objectives be translated into context and rewards? In some scenarios, such as
-request routing for load balancing, the translation is easy. In others, like
-replica placement, the translation is hard.
+Thus we have reduced our concerns to generating an appropriate context,
+executing an action (told to us by the learnin system), and reporting reward
+information.  We believe that fundamental properties of distributed systems make
+even these tasks difficult, as described below.
 
 \subsection*{Locality}
+
 Decisions made by an ML agent in distributed system are likely to rely on input
@@
 
-\subsection*{Feature selection}
+\subsection*{Failures}
+
 What features of the run-time system are important to the ML agent? How can they
@@
 
-\subsection*{Training}
+\subsection*{Randomization}
+
 An ML agent can be trained offline with pre-labeled data. However, is this
@@
 
-Two drawbacks: opacity and scope.
+There are two drawbacksTwo drawbacks: opacity and scope.
 

### 2016-01-22 writeups/msr_workshop/paper.tex
locality

@@
 \documentclass[a4paper,twocolumn]{article}
+
+\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
 
@@
 
-Decisions made by an ML agent in distributed system are likely to rely on input
-data collected from diverse and physically distributed components. As a result,
-the input to the learning loop might be stale or inconsistent. What could be
-done to mitigate those issues while making reasonable real-time decisions?
-Additionally, the decisions outputted by the learning agent need to be executed
-by diverse and distributed components. How could routine distributed systems
-issues such failures and delays in an asynchronous setting affect the learning
-system?
+Decisions in a distributed system often rely on contextual information that
+resides on diverse and physically disparate components. This information may not
+be readily available at the point at which the decision is made (e.g., the load
+balancer).  As a result, the context supplied to the learning system may be
+stale. What are the effects of staleness, and can the learning system cope with
+them (e.g., by relying on ''smoothness'' properties of the context)?  Can we
+leverage recent hardware or software techniques to reduce or eliminate
+staleness?
+
+Another disparity occurs between the place a decision is made (e.g., the load
+balancer) and where the resulting action is taken (e.g., the server selected to
+process a request). This means that the enforcement of the action cannot be
+guaranteed by the decision maker, and indeed it may be altered or reversed.
+Contextual learning systems are capable of tolerating downstream logic
+that alters the decision, provided certain independence assumptions
+aren't violated and rewards are reported accurately. Are there circumstances in
+which distributed systems would violate these assumptions enough to affect
+performance?
+
+\subsection*{Failures}
+
+What features of the run-time system are important to the ML agent? How can they
+be captured effectively? Can the ML agent learn from incomplete data? For
+example, if an ML agent needs to make decisions regarding performance
+optimization, would ignoring garbage collection activities or the VMM result in
+corrupting the learned model? How sensitive should the learner be?
+
+- quesiton of what features to include is there, but being conservative usually the right
+way to go
+- bigger quesiton is how ml component deals with omissions and other problems. Separate
+into failure sections? Yeah, replace this with ``failures'' and move feature selection stuff
+down to last section
+How could routine
+distributed systems issues such failures and delays in an asynchronous setting affect the
+learning system?
 
@@
 and exploration randomization is different at each level
-
-\subsection*{Failures}
-
-What features of the run-time system are important to the ML agent? How can they
-be captured effectively? Can the ML agent learn from incomplete data? For
-example, if an ML agent needs to make decisions regarding performance
-optimization, would ignoring garbage collection activities or the VMM result in
-corrupting the learned model? How sensitive should the learner be?
-
-- quesiton of what features to include is there, but being conservative usually the right
-way to go
-- bigger quesiton is how ml component deals with omissions and other problems. Separate
-into failure sections? Yeah, replace this with ``failures'' and move feature selection stuff
-down to last section
 

### 2016-01-22 writeups/msr_workshop/paper.tex
Modifications, mainly to section 1

--- 2016-01-22 writeups/msr_workshop/paper.tex
@@
 For example, to build a simple caching layer that guarantees a certain hit rate,
-response latency, and throughput, developers need to take into account the expected
-workload hitting the cache, the number of machines available to run it, the network
-connecting these machines, the storage hierarchy within each machine, and interference
-from other applications running on the network and machines.  A change in any one
-of these factors could render the service inefficient.
+response latency, and throughput, developers need to take into account the
+expected workload accessing the cache, the number of machines available to run
+it, the network connecting these machines, the storage hierarchy within each
+machine, and interference from other applications running on the network and
+machines.  A change in any one of these factors could render the service
+woefully inefficient.
 
@@
 to build custom systems that are optimized for specific configurations and
-objective functions.  Is there a general, principled approach for dealing with
-complexity in distributed systems? Hand-tuning the design and configuration of a system
-does not scale well with increased complexity and changes in software design, hardware
-architecture, deployment environment, and service-level objectives.
+objective functions.  Is there a general, principled, approach for dealing with
+complexity in distributed systems? Hand-tuning the design and configuration of a
+system does not scale well with increased complexity or changes in software
+design, hardware architecture, deployment environment, and service-level
+objectives.
 
@@
 view a {\em decision} made by a distributed system in the framework of
-contextual online learning with partial feedback. This is a natural fit: we typically
-have plenty of context surrounding a decision (which we may not know how to use), the decision
-is made online, and we only receive feedback for the decision that was made. Thus, we can
-replace a hand-designed policy with a new policy that has been optimized by a contextual
-learning system. By supplying rich contexts to this policy, we can capture the complexity of
-our environment without having to understand it; by optimizing the policy continuously, we can
-can adapt to changes in the environment or service over time.
+contextual online learning with partial feedback. This is a natural fit for
+distributed systems where: there is plenty of context surrounding each decision,
+decisions are made online, and feedback is received only for decisions that were
+executed.  Thus, a hand-designed policy in a distributed systems can be replaced
+with policy that has been optimized by a contextual learning system. By
+supplying rich contexts to this policy, developers capture the complexity of our
+environment without having to understand it. Additionally, by optimizing the
+policy continuously, the learning system can swiftly adapt to changes in the
+environment or service.
 
-Machine learning has certainly been used in the past to optimize system decisions. Most
-of this use has been restricted to offline settings with full information: for example,
-supervised learning is often used to train models on data that has been annotated with
-correct labels. Some systems work in an online setting with full information: for example,
-a cache that applies an experts learning framework to dynamically switch between existing
-(hand-designed) cache eviction policies, by running all policies in parallel and measuring
-their performance. Systems that work in an online setting with partial information tend to
-be for news or ad recommendation, but examples from systems infrastructure exist: for example,
-caching dynamic query content [], tuning parameters of multicore data structures [], allocating
-resources to cluster computing tasks [], and allocating servers to web applications [].
-In all of these systems, the context used to make a decision, and the
-decision itself, are {\em local} to a machine. This is achieved by either devising a decentralized
-algorithm in the first place, or giving up on global optimality. Thus we are not aware of any
-system that makes ``global'' decisions based on truly distributed architecture or state.
+Machine learning has certainly been used in the past to optimize system
+decisions. Most of this use has been restricted to either offline settings or
+expect {\em full information}, whereby a supervised learner is trained on data
+that has been annotated with correct labels.
+Some systems work in an online setting with full information. An example of that
+is a caching layer that uses a learning system to dynamically switch between
+existing (hand-designed) cache eviction policies, by running all policies in
+parallel and measuring their performance.
+Systems that work in an online setting with partial information
+tend to be for news or ad recommendation, but examples from systems
+infrastructure exist. For example, caching dynamic query content [], tuning
+parameters of multicore data structures [], allocating resources to cluster
+computing tasks [], and allocating servers to web applications [].  In all of
+these systems, the context used to make a decision, and the decision itself, are
+{\em local} to a machine. This is achieved by either devising a decentralized
+algorithm in the first place, or giving up on global optimality. We are not
+aware of any system that makes ``global'' decisions based on truly distributed
+architecture or state.
 
-And yet we use hand-designed policies to make these decisions all the time, such as for request
-routing, replica placement, replica selection, cloud resource allocation, failure recovery tasks, 
-and others. What is it about distributed systems that makes it difficult to apply online learning
-at standard decision points? This proposal identifies fundamental challenges posed by distributed
-systems that must be analyzed and addressed in order to make online learning a standard tool for
-cloud optimization.  These challenges hold even if we assume that the online learning system is
-a black-box that delivers optimal decisions with zero latency--in other words,
-they are fundamental to the way we design distributed systems.  We begin by highlighting some of the unknown factors
-that contribute to these challenges, and then list the challenges themselves.
+%% Hussam: I don't know what is this "and yet" referencing. So I am taking it out.
+%%
+%%And yet we use hand-designed policies to make these decisions all the time, such
+%%as for request routing, replica placement, replica selection, cloud resource
+%%allocation, failure recovery tasks, and others.
+
+What is it about distributed systems that makes it difficult to apply online
+learning at decision points such as request routing, replica placement, cloud
+resource allocation, and failure recovery? This proposal identifies fundamental
+challenges posed by distributed systems that must be analyzed and addressed in
+order to make online learning a standard tool for distributed infrastructure
+systems.  These challenges hold even if we assume that the online learning
+system is a black-box that delivers optimal decisions with zero latency. In
+other words, they are fundamental to the way we design distributed systems.  We
+begin by highlighting some of the factors that contribute to these challenges,
+and then list the challenges themselves.
 
@@
 
-Systems often make decisions without complete knowledge of the factors that affect,
-or could affect, those decisions.  For example, a caching system evicts items based
-on what it believes future requests will look like, without actually knowing the future.
-In other cases, incomplete knowledge is due to the absence of data-computation locality.
-For example, a load balancer makes decisions based on what it thinks the current load on
-machines is, e.g. based on data it collected recently, rather than the actual load at that
-very instant in time.
+Systems often make decisions without complete knowledge of the factors that
+affect, or could affect, those decisions.  For example, a caching system evicts
+items based on what it believes future requests will look like, without actually
+knowing the future.  In other cases, incomplete knowledge is due to the absence
+of data-computation locality.  For example, a load balancer makes decisions
+based on what it thinks is the current load on other machines, e.g. based on
+data it collected recently, rather than the actual load at that very instant in
+time.
 
-In both of these situations, assumptions are made about the unknowns offline during
-\emph{design time}, resulting in different routing and caching algorithms. A change to
-the \emph{run-time} environment could easily render these choices suboptimal.
+In both of these situations, assumptions are made about the unknowns offline
+during \emph{design time}, resulting in different request routing and caching
+algorithms. A change to the \emph{run-time} environment could easily render
+these choices suboptimal.
 
@@
 In some situations, system designers have no way of knowing deployment details
-that might impact a system's performance. For example, a system could be
-built for external deployment and used by third-party customers, or it could be
-deployed in a multi-tenant environment where co-located services could
-affect its performance in unexpected ways.
-
-In these cases, it is nearly impossible for developers to account for all 
-unknown variables at design time.
+that might impact a system's performance. For example, a system could be built
+for external deployment and used by third-party customers, or it could be
+deployed in a multi-tenant environment where co-located services could affect
+its performance in unexpected ways. In these cases, it is nearly impossible for
+developers to account for all unknown variables at design time.
 
@@
 
-As mentioned earlier, we adopt the framework of contextual online learning
-with partial feedback.  To separate the systems concerns we are trying to
-highlight from machine learning concerns, we abstract the learning system as a
-black-box and assume it provides optimal decisions with zero latency. For a
-given decision over a discrete action space, the interface to the learning
-system is as follows: given a {\em context} that summarizes the state of the
-world, the learning system chooses an {\em action} to take, which our system
-does and later reports a {\em reward} indicating how good the action was. An
-example of a learning system that supports this interface is the Decision
-Service~\footnote{http://aka.ms/mwt}. An example of a distributed systems
-decision is routing requests at a load balancer: the context could be load
-information of each machine in the cluster, the action could be one of the
-possible machines, and the reward could be the latency of processin the request.
+As mentioned earlier, we adopt the framework of contextual online learning with
+partial feedback.  To separate the systems concerns we are trying to highlight
+from machine learning concerns, we abstract the learning system as a black-box
+and assume it provides optimal decisions with zero latency. For a given decision
+over a discrete action space, the interface to the learning system is as
+follows: given a {\em context} that summarizes the state of the world, the
+learning system chooses an {\em action} to take, which our system does and later
+reports a {\em reward} indicating how good the action was. An example of a
+learning system that supports this interface is the Decision
+Service~\footnote{http://aka.ms/mwt}. Now consider a load balancer using that
+learning system to make its request-routing decisions. The context for each
+decision could be load information of each machine in the cluster, the action
+could be a specific machine to route the request to, and the reward could be the
+latency of processing the request.
 
 Thus we have reduced our concerns to generating an appropriate context,
-executing an action (told to us by the learnin system), and reporting reward
+executing an action (told to us by the learning system), and reporting reward
 information.  We believe that fundamental properties of distributed systems make
@@
     other?
+   \item Should feature vectors cut across the levels or not?
 \end{itemize}
 
-- Should feature vectors cut across the levels or not?
 - In general should be composable if at each level can define decision and reward,

### 2016-01-22 writeups/msr_workshop/paper.tex
renamed 'full information' to 'full feedback', and defined what partial feedback means

@@
 view a {\em decision} made by a distributed system in the framework of
-contextual online learning with partial feedback. This is a natural fit for
-distributed systems where: there is plenty of context surrounding each decision,
-decisions are made online, and feedback is received only for decisions that were
-executed.  Thus, a hand-designed policy in a distributed systems can be replaced
-with policy that has been optimized by a contextual learning system. By
-supplying rich contexts to this policy, developers capture the complexity of our
-environment without having to understand it. Additionally, by optimizing the
-policy continuously, the learning system can swiftly adapt to changes in the
-environment or service.
+contextual online learning with {\em partial feedback} (where the learner only
+gets to observe the result of the action it took, rather than all possible
+actions). This is a natural fit for distributed systems where: there is plenty
+of context surrounding each decision, decisions are made online, and feedback is
+received only for decisions that were executed.  Thus, a hand-designed policy in
+a distributed systems can be replaced with policy that has been optimized by a
+contextual learning system. By supplying rich contexts to this policy,
+developers capture the complexity of our environment without having to
+understand it. Additionally, by optimizing the policy continuously, the learning
+system can swiftly adapt to changes in the environment or service.
+
+%% Hussam: I changed "full information" to "full feedback" to stay consistent
+%% with the naming above. Confirmed with Akshay that this is correct terminology
 
@@
 decisions. Most of this use has been restricted to either offline settings or
-expect {\em full information}, whereby a supervised learner is trained on data
+expect {\em full feedback}, whereby a supervised learner is trained on data
 that has been annotated with correct labels.
-Some systems work in an online setting with full information. An example of that
+Some systems work in an online setting with full feedback. An example of that
 is a caching layer that uses a learning system to dynamically switch between
@@
 parallel and measuring their performance.
-Systems that work in an online setting with partial information
+Systems that work in an online setting with partial feedback
 tend to be for news or ad recommendation, but examples from systems

### 2016-01-22 writeups/msr_workshop/paper.tex
more content

--- 2016-01-22 writeups/msr_workshop/paper.tex
@@
 
+While reward information is also not available at the decision location, this is
+the norm in contextual learning settings, where reward information is only
+discernible after a delay and often by a separate subsystem. Thus, we do not
+expect this to pose a problem. (The faster reward information is collected, the
+faster the resulting datapoint can be used by the learning subsystem.)
+
 \subsection*{Failures}
 
-What features of the run-time system are important to the ML agent? How can they
-be captured effectively? Can the ML agent learn from incomplete data? For
-example, if an ML agent needs to make decisions regarding performance
-optimization, would ignoring garbage collection activities or the VMM result in
-corrupting the learned model? How sensitive should the learner be?
+The disparity in locality above results in a lack of fate sharing between the
+distributed system components involved in a decision.  Failures in any of
+htese components have implications on the inputs to the learning system. For
+example, server failures in a load balanced cluster could mean that certain
+context information is staler than others or unavailable altogether. How do we
+convey this ''lack of information'' to the learning system, and how does it
+cope with sporadically incomplete data? Should failures be encoded in the
+context somehow?
 
-- quesiton of what features to include is there, but being conservative usually the right
-way to go
-- bigger quesiton is how ml component deals with omissions and other problems. Separate
-into failure sections? Yeah, replace this with ``failures'' and move feature selection stuff
-down to last section
-How could routine
-distributed systems issues such failures and delays in an asynchronous setting affect the
-learning system?
+Failures in distributed systems are hard to distinguish from asynchrony in some
+cases. What other ways does asynchrony affect the learning system?
+
+There is another failure mode in which important contextual information is
+simply not captured due to designer ommision. In general contextual learning
+algorithms scale very well with the size of the size of the context, so it is
+generally a good idea to include more data than necessary. In some cases,
+however, it may not be easy to access this data, e.g. obtaining statistics
+about a background garbage collection process or about a VMM.
 
 \subsection*{Hierarchy}
+
 Distributed systems are often designed in a hierarchical manner, where
-subcomponents are monitoring and managing others. These components and layers of
-indirection simplify reasoning about the system at large. However, it is not
-clear how hierarchy will affect an ML agent.
+subcomponents monitor and manage others. These components and layers of
+indirection simplify reasoning, building, and maintaining the system at
+large. However, it is not clear how hierarchy should be handled with respect to
+optimizing decisions via contextual learning. In general contextual learning is
+composable, in that decisions of subcomponents can be used to influence
+higher-level components, downstream components can make decisions influenced by
+the decisions of upstream components, and so on.  Should the context for each
+decision be limited to the level it is at in the hierarchy, or is it better for
+contexts to span multiple (or all) levels? Should decision points roughly mimic
+the hierarchy of the system? Consider deep memory hierarchy, for example. One
+could imagine that by capturing contextual information across the entire
+hierarchy, one could decide at what level to cache a piece of data, instead of
+only deciding on the current level and relying on evictions to move between
+levels.
+
+\ignore{
 \begin{itemize}
@@
 \end{itemize}
-
-- Should feature vectors cut across the levels or not?
-- In general should be composable if at each level can define decision and reward,
-and exploration randomization is different at each level
+}
 
@@
 
+Contextual learning uses exploration (controlled randomization) to discover
+potentially good strategies. This is the famous explore-exploit tradeoff and is
+well established to be necessary for proper online optimization. Randomization
+can be encapsulated and handled by the learning system, but there are two
+situations in which it might emerge to the systems level. First, if multiple
+components are making decisions independently, it is important to ensure that
+thre randomization used by each (e.g. the seed passed to the learning system)
+are independent of each other. Second, many distributed systems already use
+randomization inherently, such as randomized load balancing or randomized cache
+eviction strategies. Can this randomization be leveraged by the learning
+system to achieve its exploration goals naturally?
+
+\ignore{
 An ML agent can be trained offline with pre-labeled data. However, is this
@@
 exploration vs exploitation of learned strategies?
+}
 
-- Probably omit this, or replace with ``Randomization''. Basically, we already know
-that exploratoin-exploitation is the right approach, and that will be encapsulated
-by the ML component (its decisions will reflect this, and it will take care of
-computing probabilities etc.). But there's an interesting point about inherent randomization
-in systems
-
+\ignore{
 \section{Expanding the scope}
@@
 asudden changes the weight on other features.
+}
 

### 2016-01-22 writeups/msr_workshop/paper.tex
more content

@@
 
-\section{Using Online Learning in Distributed Systems}
+\section{Challenges}
 
@@
 
+\section{Drawbacks}
+
+The approach we have outlined above has at least two drawbacks. One is that we
+are using machine-learned models to capture the complexity of a distributed
+systems decision. This means that the resulting optimal policy will be opaque to
+us, making it difficult to discover what contextual information matters, for
+example.  We are independently pursuing work in cracking models open to see what
+can be learned from them, but in general this is not always possible nor is it
+reliable to draw conclusion from things like learned weights. In effect, our
+position is to accept this opacity as the cost of offloading the complexity of
+making the right decision to the learning system.
+
+The second drawback is that our proposal only replaces hand-designed policies at
+existing decision points. In other words, machine learning is used as an
+alternative policy within an existing design and architecture. Can machine
+learning influence the design/architecture of the system itself? Systems like
+the Decision Service advocate the former approach, but it is conceivable that in
+the future distributed systems will be co-designed based on learning objectives.
+
 \ignore{
-\section{Expanding the scope}
-
-There are two drawbacksTwo drawbacks: opacity and scope.
-
-Opacity (also, which features matter?)
-- We are also pursuing work on cracking models open to see what they are doing inside.
+We are also pursuing work on cracking models open to see what they are doing inside.
 There are different ways to gain insight as to what a model is doing. For example a simple

### 2016-01-22 writeups/msr_workshop/paper.tex
Section 2 mods

--- 2016-01-22 writeups/msr_workshop/paper.tex
@@
 by diverse and distributed components. How could routine distributed systems
-issues such failures and delays in an asynchronous setting affect the learning
+issues such as message delays in an asynchronous setting affect the learning
 system?
@@
 
-- In general should be composable if at each level can define decision and reward,
-and exploration randomization is different at each level
+%% Sid:
+%% In general should be composable if at each level can define decision and reward,
+%% and exploration randomization is different at each level
 
@@
 
-What features of the run-time system are important to the ML agent? How can they
-be captured effectively? Can the ML agent learn from incomplete data? For
-example, if an ML agent needs to make decisions regarding performance
-optimization, would ignoring garbage collection activities or the VMM result in
-corrupting the learned model? How sensitive should the learner be?
+Failures are an expected occurrence in every distributed system. How do learning
+systems cope with data omission or inability to execute decisions due to
+failures? Can the ML agent learn from incomplete or potentially corrupt data?
+How sensitive should the learner be to transient conditions in the system?
 
-- quesiton of what features to include is there, but being conservative usually the right
-way to go
-- bigger quesiton is how ml component deals with omissions and other problems. Separate
-into failure sections? Yeah, replace this with ``failures'' and move feature selection stuff
-down to last section
+%% What features of the run-time system are important to the ML agent? How can they
+%% be captured effectively? Can the ML agent learn from incomplete data? For
+%% example, if an ML agent needs to make decisions regarding performance
+%% optimization, would ignoring garbage collection activities or the VMM result in
+%% corrupting the learned model? How sensitive should the learner be?
+%% 
+%% - quesiton of what features to include is there, but being conservative usually the right
+%% way to go
+%% - bigger quesiton is how ml component deals with omissions and other problems. Separate
+%% into failure sections? Yeah, replace this with ``failures'' and move feature selection stuff
+%% down to last section
 
-\subsection*{Randomization}
-
-An ML agent can be trained offline with pre-labeled data. However, is this
-possible in the context of distributed systems? An online learning systems is
-more logical in this context, however, how do we quantify the costs of
-exploration vs exploitation of learned strategies?
-
-- Probably omit this, or replace with ``Randomization''. Basically, we already know
-that exploratoin-exploitation is the right approach, and that will be encapsulated
-by the ML component (its decisions will reflect this, and it will take care of
-computing probabilities etc.). But there's an interesting point about inherent randomization
-in systems
+%% Hussam: Sid, I don't understand what you mean by randomization in this context
+%% So I am taking this subsection out.
+%% \subsection*{Randomization}
+%% 
+%% An ML agent can be trained offline with pre-labeled data. However, is this
+%% possible in the context of distributed systems? An online learning systems is
+%% more logical in this context, however, how do we quantify the costs of
+%% exploration vs exploitation of learned strategies?
+%% 
+%% - Probably omit this, or replace with ``Randomization''. Basically, we already know
+%% that exploratoin-exploitation is the right approach, and that will be encapsulated
+%% by the ML component (its decisions will reflect this, and it will take care of
+%% computing probabilities etc.). But there's an interesting point about inherent randomization
+%% in systems
+%%
 

### 2016-01-22 writeups/msr_workshop/paper.tex

@@
 
-Decisions made by an ML agent in distributed system are likely to rely on input
-data collected from diverse and physically distributed components. As a result,
-the input to the learning loop might be stale or inconsistent. What could be
-done to mitigate those issues while making reasonable real-time decisions?
-Additionally, the decisions outputted by the learning agent need to be executed
-by diverse and distributed components. How could routine distributed systems
-issues such as message delays in an asynchronous setting affect the learning
-system?
+Decisions in a distributed system often rely on contextual information that
+resides on diverse and physically disparate components. This information may not
+be readily available at the point at which the decision is made (e.g., the load
+balancer).  As a result, the context supplied to the learning system may be
+stale. What are the effects of staleness, and can the learning system cope with
+them (e.g., by relying on ''smoothness'' properties of the context)?  Can we
+leverage recent hardware or software techniques to reduce or eliminate
+staleness?
 
-\subsection*{Hierarchy}
-Distributed systems are often designed in a hierarchical manner, where
-subcomponents are monitoring and managing others. These components and layers of
-indirection simplify reasoning about the system at large. However, it is not
-clear how hierarchy will affect an ML agent.
-\begin{itemize}
-  \item Will hierarchy aid the learning system via parallelism? Or will it hinder
-    it since not all features are visible at every level?
-  \item Should learning happen in each component of the system? Or should it be
-    centralized?
-  \item If learning is happening at multiple levels of the hierarchy, will that
-    create interference between the learners? Or will they complement each
-    other?
-   \item Should feature vectors cut across the levels or not?
-\end{itemize}
+Another disparity occurs between the place a decision is made (e.g., the load
+balancer) and where the resulting action is taken (e.g., the server selected to
+process a request). This means that the enforcement of the action cannot be
+guaranteed by the decision maker, and indeed it may be altered or reversed.
+Contextual learning systems are capable of tolerating downstream logic
+that alters the decision, provided certain independence assumptions
+aren't violated and rewards are reported accurately. Are there circumstances in
+which distributed systems would violate these assumptions enough to affect
+performance?
 
-%% Sid:
-%% In general should be composable if at each level can define decision and reward,
-%% and exploration randomization is different at each level
+While reward information is also not available at the decision location, this is
+the norm in contextual learning settings, where reward information is only
+discernible after a delay and often by a separate subsystem. Thus, we do not
+expect this to pose a problem. (The faster reward information is collected, the
+faster the resulting data point can be used by the learning subsystem.)
 
@@
 
-Failures are an expected occurrence in every distributed system. How do learning
+The disparity in locality above results in a lack of fate sharing between the
+distributed system components involved in a decision.  Failures in any of
+these components have implications on the inputs to the learning system. For
+example, server failures in a load balanced cluster could mean that certain
+context information is staler than others or unavailable altogether. How do we
+convey this ''lack of information'' to the learning system, and how does it
+cope with sporadically incomplete data? Should failures be encoded in the
+context somehow?
+How do learning
 systems cope with data omission or inability to execute decisions due to
@@
 
-%% What features of the run-time system are important to the ML agent? How can they
-%% be captured effectively? Can the ML agent learn from incomplete data? For
-%% example, if an ML agent needs to make decisions regarding performance
-%% optimization, would ignoring garbage collection activities or the VMM result in
-%% corrupting the learned model? How sensitive should the learner be?
-%% 
-%% - quesiton of what features to include is there, but being conservative usually the right
-%% way to go
-%% - bigger quesiton is how ml component deals with omissions and other problems. Separate
-%% into failure sections? Yeah, replace this with ``failures'' and move feature selection stuff
-%% down to last section
+\subsection*{Hierarchy}
 
-%% Hussam: Sid, I don't understand what you mean by randomization in this context
-%% So I am taking this subsection out.
-%% \subsection*{Randomization}
-%% 
-%% An ML agent can be trained offline with pre-labeled data. However, is this
-%% possible in the context of distributed systems? An online learning systems is
-%% more logical in this context, however, how do we quantify the costs of
-%% exploration vs exploitation of learned strategies?
-%% 
-%% - Probably omit this, or replace with ``Randomization''. Basically, we already know
-%% that exploratoin-exploitation is the right approach, and that will be encapsulated
-%% by the ML component (its decisions will reflect this, and it will take care of
-%% computing probabilities etc.). But there's an interesting point about inherent randomization
-%% in systems
-%%
+Distributed systems are often designed in a hierarchical manner, where
+subcomponents monitor and manage others. These components and layers of
+indirection simplify reasoning, building, and maintaining the system at
+large. However, it is not clear how hierarchy should be handled with respect to
+optimizing decisions via contextual learning. In general contextual learning is
+composable, in that decisions of subcomponents can be used to influence
+higher-level components, downstream components can make decisions influenced by
+the decisions of upstream components, and so on.  Should the context for each
+decision be limited to the level it is at in the hierarchy, or is it better for
+contexts to span multiple (or all) levels? Should decision points roughly mimic
+the hierarchy of the system? Consider deep memory hierarchy, for example. One
+could imagine that by capturing contextual information across the entire
+hierarchy, one could decide at what level to cache a piece of data, instead of
+only deciding on the current level and relying on evictions to move between
+levels.
 
-\section{Expanding the scope}
+\subsection*{Randomization}
 
-There are two drawbacksTwo drawbacks: opacity and scope.
-
-Opacity (also, which features matter?)
-- We are also pursuing work on cracking models open to see what they are doing inside.
-There are different ways to gain insight as to what a model is doing. For example a simple
-linear model may learn a vector of weights on the supplied features. These weights can be
-studied, and indeed some techniques such as lasso are oftne used to reduce the number of features. However, these
-techniques need to be handled carefully. Just becasue a feature has 0 weight doesn't mean it
-can be removed: perhaps it is unimportant preceisely because it exists, and its removal all of
-asudden changes the weight on other features.
+Contextual learning uses exploration (controlled randomization) to discover
+potentially good strategies. This is the famous explore-exploit trade-off and is
+well established to be necessary for proper online optimization. Randomization
+can be encapsulated and handled by the learning system, but there are two
+situations in which it might emerge to the systems level. First, if multiple
+components are making decisions independently, it is important to ensure that
+the randomization used by each (e.g. the seed passed to the learning system)
+are independent of each other. Second, many distributed systems already use
+randomization inherently, such as randomized load balancing or randomized cache
+eviction strategies. Can this randomization be leveraged by the learning
+system to achieve its exploration goals naturally?
 
 \ignore{
-contextual multi-armed bandits. idea is to provide a context, get an action,
-and only observe reward for that action. This matches the dist systems world well. It will generate new
-policies using ML that we could not come up with by just thinking, and may not
-even be able to describe. We argue in favor of opacity and rely instead on the guarantees
-provided by the ML optimization.
+An ML agent can be trained offline with pre-labeled data. However, is this
+possible in the context of distributed systems? An online learning systems is
+more logical in this context, however, how do we quantify the costs of
+exploration vs exploitation of learned strategies?
 }
 
-Scope:
-The above replaces the policy behind a decision with one backed by an ML learning system.
-Indeed systems like hte DS were designed to intervene at the decision making point in a
-slim way. The above outlines the issues which tackled would result in ML based policies,
-new ones. However, once the above has been mastered, there is the question of whether ML
-can influence distributed systems design at a higher level. For example, can it suggest
-different mechanisms for achieving a goal, or perhaps a different design of the system in the first place?
+\section{Drawbacks}
+
+The approach we have outlined above has at least two drawbacks. One is that we
+are using machine-learned models to capture the complexity of a distributed
+systems decision. This means that the resulting optimal policy will be opaque to
+us, making it difficult to discover what contextual information matters, for
+example.  We are independently pursuing work in cracking models open to see what
+can be learned from them, but in general this is not always possible nor is it
+reliable to draw conclusion from things like learned weights. In effect, our
+position is to accept this opacity as the cost of offloading the complexity of
+making the right decision to the learning system.
+
+The second drawback is that our proposal only replaces hand-designed policies at
+existing decision points. In other words, machine learning is used as an
+alternative policy within an existing design and architecture. Can machine
+learning influence the design/architecture of the system itself? Systems like
+the Decision Service advocate the former approach, but it is conceivable that in
+the future distributed systems will be co-designed based on learning objectives.
+
+%% Scope:
+%% The above replaces the policy behind a decision with one backed by an ML learning system.
+%% Indeed systems like hte DS were designed to intervene at the decision making point in a
+%% slim way. The above outlines the issues which tackled would result in ML based policies,
+%% new ones. However, once the above has been mastered, there is the question of whether ML
+%% can influence distributed systems design at a higher level. For example, can it suggest
+%% different mechanisms for achieving a goal, or perhaps a different design of the system in the first place?
 

### 2016-01-24 writeups/msr_workshop/paper.tex
completed and cleaned up

@@
-\documentclass[a4paper,twocolumn]{article}
-
+\documentclass[letterpaper,twocolumn]{article}
+%\documentclass[pdftex,twocolumn,10pt,letterpaper]{article}
 \usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
@@
 \title{
-  It's Never Too Late To Learn\\
+  It's Never Too Late to Learn:\\
   \emph{Using Online Machine Learning in Distributed Systems}
 }
-\author{}
+\author{Hussam Abu-Libdeh and Siddhartha Sen \\ \small{Microsoft Research NYC}}
 \date{}
@@
 as the service evolves with time to accommodate hardware/environment changes or
-support new features.
+support new features. Some of the unknown variables are known: a caching layer
+must cope with changing workload distributions; a load balancer might have stale
+load information from the servers. But some of the unknowns are unknown: a
+system could be deployed in a multi-tenant environment where co-located
+services affect its performance in unexpected ways.
+
+\ignore{
 For example, to build a simple caching layer that guarantees a certain hit rate,
@@
 woefully inefficient.
+}
 
@@
 complexity in distributed systems? Hand-tuning the design and configuration of a
-system does not scale well with increased complexity or changes in software
-design, hardware architecture, deployment environment, and service-level
-objectives.
+system does not scale well with increased complexity or changes in the
+environment or service-level objectives.
 
 We propose to use online learning to cope with this complexity. Specifically, we
-view a {\em decision} made by a distributed system in the framework of
-contextual online learning with {\em partial feedback} (where the learner only
-gets to observe the result of the action it took, rather than all possible
-actions). This is a natural fit for distributed systems where: there is plenty
-of context surrounding each decision, decisions are made online, and feedback is
-received only for decisions that were executed.  Thus, a hand-designed policy in
-a distributed systems can be replaced with policy that has been optimized by a
-contextual learning system. By supplying rich contexts to this policy,
-developers capture the complexity of our environment without having to
-understand it. Additionally, by optimizing the policy continuously, the learning
-system can swiftly adapt to changes in the environment or service.
+cast the decisions made by a distributed system in the framework of
+{\em contextual online learning with partial feedback}.  This is a natural fit
+for many decisions where: there is plenty of context relevant to the decision, the
+decision is made online, and feedback is only received for the action that
+was taken.  Thus, we can replace a hand-designed policy for the decision with a
+policy that has been optimized by a contextual learning system. By supplying
+rich contexts to the learner, developers can capture the complexity of the
+environment without having to understand it.  By training the learner
+continuously, the system can adapt swiftly to changes in the environment
+or service.
 
@@
 
+Online learning has been used in systems before. Some of these systems require
+full feedback, which is not practical, but others operate in the partial
+feedback setting.  The latter systems tend to be recommendation systems (e.g.,
+for personalized news or ads), but some infrastructure systems exist, such as
+for caching dynamic query content or allocating resources to
+cluster jobs.  In all of these systems, the contextual decisions are made {\em
+locally} within a machine, either because the system is not distributed, or it
+uses a decentralized algorithm, or it does not aim for global optimality.
+\ignore{We are not aware of any system that makes ``global'' decisions based on truly
+distributed architecture or state.}
+
+\ignore{
 Machine learning has certainly been used in the past to optimize system
@@
 architecture or state.
+}
 
@@
 
-What is it about distributed systems that makes it difficult to apply online
-learning at decision points such as request routing, replica placement, cloud
-resource allocation, and failure recovery? This proposal identifies fundamental
+More importantly, prior work does not abstract, analyze, or address the
+fundamental challenges imposed by distributed systems when applying online
+learning.  The purpose of this proposal is to identify these challenges and
+motivate us to start working on them.  We believe these are systems
+challenges: that is, they hold even if we assume black-box access to a
+learning system that provides optimal decisions in real-time. By overcoming
+these challenges, we stand to gain principled mechanisms and tools for
+automating intelligence in our distributed infrastructure systems, such as our
+cloud.
+
+\ignore{
+and points such as request routing, replica placement,
+cloud resource allocation, and failure recovery? This proposal identifies fundamental
 challenges posed by distributed systems that must be analyzed and addressed in
@@
 and then list the challenges themselves.
+}
 
+\ignore{
 \subsection*{Known unknowns}
@@
 }
+}
 
@@
 
-As mentioned earlier, we adopt the framework of contextual online learning with
-partial feedback.  To separate the systems concerns we are trying to highlight
+\ignore{We consider classic decisions made by distributed systems where the
+choice of action is discrete, such as: where to route a request, how many replicas to
+place (and where), or when to raise an alarm during failures.}
+
+For a given decision, we can view the world as having two components: the
+distributed system and the learning system, which we treat as a black box.  The
+system interacts with the learner as follows: it provides a {\em context}
+to the learner summarizing the state of the world, the learner returns 
+an {\em action} to take, the system executes the action and reports a {\em
+reward} to the learner indicating how good it was.  An example of a
+contextual learning system that supports this interface is the Decision
+Service\footnote{http://aka.ms/mwt}.  An example of a decision is which
+server a load balancer should send a request to: the context could be the load
+information of every server, the action is the chosen server, and
+the reward could be the latency of the request.
+
+\ignore{
+We separate the world into a As mentioned earlier, we
+adopt the framework of contextual online learning with partial feedback.  To separate the systems concerns we are trying to highlight
 from machine learning concerns, we abstract the learning system as a black-box
@@
 latency of processing the request.
+}
 
-Thus we have reduced our concerns to generating an appropriate context,
-executing an action (told to us by the learning system), and reporting reward
-information.  We believe that fundamental properties of distributed systems make
-even these tasks difficult, as described below.
+Thus the system is responsible for generating an appropriate context, executing
+an action, and reporting the reward.  The following properties of distributed
+systems make these tasks challenging (there may be others).
 
-\subsection*{Locality}
+\paragraph{Locality.}
+Decisions in a distributed system may rely on information that resides on
+diverse and physically disparate components. This information may not be readily
+available at the point of decision (e.g., the load balancer).
+As a result, the context supplied to the learner may be stale or
+incomplete.  How does staleness affect the learner's ability to optimize
+decisions? Can we exploit ``smoothness'' in the context to mitigate this effect?
+Can we eliminate staleness using hardware or software techniques?
 
-Decisions in a distributed system often rely on contextual information that
-resides on diverse and physically disparate components. This information may not
-be readily available at the point at which the decision is made (e.g., the load
-balancer).  As a result, the context supplied to the learning system may be
-stale. What are the effects of staleness, and can the learning system cope with
-them (e.g., by relying on ''smoothness'' properties of the context)?  Can we
-leverage recent hardware or software techniques to reduce or eliminate
-staleness?
+Another disparity may occur between the point of decision and the
+point of execution (e.g., the server chosen to handle a request).  This
+separation prevents the decider from enforcing the action; indeed, the action
+may be changed or ignored.  Contextual learning systems can tolerate
+downstream logic that alters the decision, provided rewards are
+correctly reported and certain independence assumptions are not
+violated.  Are there distributed systems that violate these assumptions?
 
-Another disparity occurs between the place a decision is made (e.g., the load
-balancer) and where the resulting action is taken (e.g., the server selected to
-process a request). This means that the enforcement of the action cannot be
-guaranteed by the decision maker, and indeed it may be altered or reversed.
-Contextual learning systems are capable of tolerating downstream logic
-that alters the decision, provided certain independence assumptions
-aren't violated and rewards are reported accurately. Are there circumstances in
-which distributed systems would violate these assumptions enough to affect
-performance?
+It is common for reward information to be reported by a separate
+component. We do not expect this to pose a problem beyond the fact that delays 
+in reward reporting cause delays in learning.
 
-While reward information is also not available at the decision location, this is
-the norm in contextual learning settings, where reward information is only
-discernible after a delay and often by a separate subsystem. Thus, we do not
-expect this to pose a problem. (The faster reward information is collected, the
+\ignore{typically reported separately by a asynchronously While reward
+information is also not available at the decision location, this is the norm in contextual learning
+settings, where reward information is only discernible after a delay and often
+by a separate subsystem. Thus, we do not expect this to pose a problem. (The faster reward information is collected, the
 faster the resulting data point can be used by the learning subsystem.)
+}
 
-\subsection*{Failures}
+\paragraph{Failures.}
+The absence of locality results in a lack of fate sharing
+between the components involved in a decision.  Failures in any of
+these components affect the inputs to the learning system.
+For example, server failures in the load balancing example could
+result in variable staleness or even missing/corrupt information in
+the context.  How should information that is known to be missing 
+be conveyed to the learner?  Can corrupt data derail the learner's optimization?
 
-The disparity in locality above results in a lack of fate sharing between the
-distributed system components involved in a decision.  Failures in any of
-these components have implications on the inputs to the learning system. For
-example, server failures in a load balanced cluster could mean that certain
-context information is staler than others or unavailable altogether. How do we
-convey this ''lack of information'' to the learning system, and how does it
-cope with sporadically incomplete data? Should failures be encoded in the
-context somehow?
-How do learning
-systems cope with data omission or inability to execute decisions due to
-failures? Can the ML agent learn from incomplete or potentially corrupt data?
-How sensitive should the learner be to transient conditions in the system?
+Failures can also prevent a component from executing the action, causing the
+system to change the action or ignore it.  Can such failures violate the
+independence assumptions mentioned above?  In general, how sensitive should
+the learner be to transient conditions in the system?
 
-\subsection*{Hierarchy}
+\paragraph{Hierarchy.}
+Components in a distributed system often interact with and manage other
+components. This hierarchy simplifies building, maintaining, and reasoning
+about the system at large. It is not clear how to take hierarchy into
+account when optimizing decisions via contextual learning.  Contextual learning
+is generally composable, in that decisions of lower-level
+components can safely influence those of higher-level components, and those of
+upstream components can influence those of downstream components.   Should the
+context for each decision be limited to its level in the hierarchy, or should it span
+multiple (or all) levels?  Should the decision points roughly mimic the
+hierarchy of the system, or should they cut across? Consider the example of a
+deep memory hierarchy. One could imagine capturing contextual
+information across the entire hierarchy and using it to decide which level to cache an item in, instead of deciding at each level in isolation and relying on evictions to move between levels.
 
-Distributed systems are often designed in a hierarchical manner, where
-subcomponents monitor and manage others. These components and layers of
-indirection simplify reasoning, building, and maintaining the system at
-large. However, it is not clear how hierarchy should be handled with respect to
-optimizing decisions via contextual learning. In general contextual learning is
-composable, in that decisions of subcomponents can be used to influence
-higher-level components, downstream components can make decisions influenced by
-the decisions of upstream components, and so on.  Should the context for each
-decision be limited to the level it is at in the hierarchy, or is it better for
-contexts to span multiple (or all) levels? Should decision points roughly mimic
-the hierarchy of the system? Consider deep memory hierarchy, for example. One
-could imagine that by capturing contextual information across the entire
-hierarchy, one could decide at what level to cache a piece of data, instead of
-only deciding on the current level and relying on evictions to move between
-levels.
+\paragraph*{Randomization.}
 
-\subsection*{Randomization}
+Contextual learning balances exploration (a type of controlled randomization)
+with exploitation so that it can discover potentially better policies
+without compromising current performance.  Exploration can be completely
+encapsulated by the learning system and manifested through the actions it
+recommends.  However, many distributed systems already use
+randomization, such as for cache eviction, load balancing, 
+and replica placement.  Can this randomization be leveraged by the learning
+system to achieve its exploration goals at no additional cost?
 
-Contextual learning uses exploration (controlled randomization) to discover
-potentially good strategies. This is the famous explore-exploit trade-off and is
-well established to be necessary for proper online optimization. Randomization
-can be encapsulated and handled by the learning system, but there are two
-situations in which it might emerge to the systems level. First, if multiple
-components are making decisions independently, it is important to ensure that
-the randomization used by each (e.g. the seed passed to the learning system)
-are independent of each other. Second, many distributed systems already use
-randomization inherently, such as randomized load balancing or randomized cache
-eviction strategies. Can this randomization be leveraged by the learning
-system to achieve its exploration goals naturally?
+\ignore{First, if multiple components are making decisions independently, it is
+important to ensure that the randomization used by each (e.g. the seed passed to
+the learning system) are independent of each other.}
 
@@
 
-The approach we have outlined above has at least two drawbacks. One is that we
-are using machine-learned models to capture the complexity of a distributed
-systems decision. This means that the resulting optimal policy will be opaque to
-us, making it difficult to discover what contextual information matters, for
-example.  We are independently pursuing work in cracking models open to see what
-can be learned from them, but in general this is not always possible nor is it
-reliable to draw conclusion from things like learned weights. In effect, our
-position is to accept this opacity as the cost of offloading the complexity of
-making the right decision to the learning system.
+The approach we have outlined has at least two drawbacks that merit their own
+investigation.  The first is that we are using machine-learned policies that are
+generally opaque, in that we cannot easily inspect them to learn things like
+which parts of the context really matter.  Cracking open such policies is
+certainly an interesting research direction, but our current view is to accept
+opacity as the price we pay for offloading complexity to the learner.
 
-The second drawback is that our proposal only replaces hand-designed policies at
-existing decision points. In other words, machine learning is used as an
-alternative policy within an existing design and architecture. Can machine
-learning influence the design/architecture of the system itself? Systems like
-the Decision Service advocate the former approach, but it is conceivable that in
-the future distributed systems will be co-designed based on learning objectives.
+The second drawback is that these policies replace hand-designed policies only
+at existing decision points.  In other words, machine learning is used as an
+alternative within an existing design and architecture, as opposed to 
+influencing the design/architecture of the system itself.  It is very
+conceivable that in the future distributed systems will be co-designed with
+learners.
 
